Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system with plausible mechanism

### 2024-07-31 _tex/index.tex
Built site for gh-pages

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system with plausible mechanism},
  pdfauthor={Jose Manuel Rivera Espejo; Tine van van Daal; Sven De De Maeyer; Steven Gillis},
  pdfkeywords={comparative judgement, directed acycilc graph, causal
analysis, probabilistic statistics},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{abstract}
Dichotomous Comparative Judgment (DCJ) requires judges to compare pairs
of stimuli to determine which one exhibits a higher degree of a specific
trait. DCJ has proven effective and reliable across various fields
\citep{Pollitt_2012b, Jones_2015, vanDaal_et_al_2016, Bartholomew_et_al_2018, Lesterhuis_2018, Bartholomew_et_al_2020, Marshall_et_al_2020, Boonen_et_al_2020}.
However, despite the method's widespread use, existing literature lacks
a clear explanation of the complexities and assumptions underpinning the
DCJ system, as well as the plausible mechanisms through which DCJ data
could be generated. This study addresses these issues by representing
DCJ within the framework of causal inference. Specifically, utilizing
the structural approach, the study develops a scientific model to
clarify plausible causal assumptions and mechanisms inherent in the DCJ
system. The study then translates this model into a probabilistic
statistical framework to estimate statistical relationships and infer
causal effects within the system. This research provides a robust
probabilistic foundation for the statistical analysis of DCJ data,
building upon Thurstone's law of comparative judgment
\citeyearpar{Thurstone_1927}. Its findings offer valuable insights for
researchers and analysts designing and implementing DCJ experiments.
\end{abstract}

\begin{keyword}
    comparative judgement \sep directed acycilc graph \sep causal
analysis \sep 
    probabilistic statistics
\end{keyword}
\end{frontmatter}

@@

@@

Regardless of complexity, DAGs can represent various causal structures
using only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. Each panel of Figure~\ref{fig-dags}
illustrates these building blocks. Figure~\ref{fig-dag_bb1} depicts two
unconnected nodes, representing an scenario where variables \(X_{1}\)
and \(X_{2}\) are not causally related. Figure~\ref{fig-dag_bb2} shows
two connected nodes, illustrating a scenario where a parent node
\(X_{1}\) exerts a causal influence on a child node \(X_{2}\).
Consequently, \(X_{2}\) is considered a \emph{descendant} of \(X_{1}\).
Figure~\ref{fig-dag_bb3} depicts a \emph{chain} (or \emph{pipe}), where
\(X_{1}\) influences \(X_{2}\), and \(X_{2}\) influences \(X_{3}\). In
this configuration, \(X_{1}\) is a parent node of \(X_{2}\), and
\(X_{2}\) a parent node of \(X_{3}\). Furthermore, the DAG shows that
\(X_{1}\) is an \emph{ancestor} of \(X_{3}\), and that the relationship
between these variables is entirely mediated by \(X_{2}\).
Figure~\ref{fig-dag_bb4} illustrates a \emph{fork}, where variables
\(X_{1}\) and \(X_{3}\) are both influenced by \(X_{2}\). In this
scenario \(X_{2}\) is a parent node of both \(X_{1}\) and \(X_{3}\).
Finally, Figure~\ref{fig-dag_bb5} depicts a \emph{collider}, also known
as \emph{inmorality}, where variables \(X_{1}\) and \(X_{3}\) are
concurrent causes of \(X_{2}\). In this configuration, \(X_{1}\) and
\(X_{3}\) are not causally related to each other but both influence
\(X_{2}\).

@@

Given the heuristic nature of DAGs, the use of fundamental building
block to construct a causal structure of a system is easier to
understand using a motivating example. The motivating example can also
serve to signal about other conventions when using DAGs. Consider a
system where variables \(X\) and \(Z\) influence a third variable \(Y\).
In this system, it is assumed that \(X\) and \(Z\) are dependent on
their own processes and are, therefore, independent from each other.
Figure~\ref{fig-dag_example1} presents the plausible causal structure of
this system. The DAG shows the endogenous variables \(V=\{X,Z,Y\}\) as
circled black nodes, indicating these variables are observed. The arrows
connecting the variables indicate the direction of causal influence,
while a lack of influence is indicated by the absence of arrows.
Moreover, the exogenous variables \(E=\{e_{X},e_{Z},e_{Y}\}\) represent
everything else that is chosen not to be modeled explicitly. These
exogenous variables, or \emph{disturbances}, are usually represented by
open circles to indicate their unobserved nature. Although this DAG
explicitly shows the exogenous variables, conventionally these are
omitted for brevity, resulting in an equivalent graph as shown in
Figure~\ref{fig-dag_example2}.

@@

\includegraphics[width=1\textwidth,height=\textheight]{images/figures/dag_example1.png}

@@

\subcaption{\label{fig-dag_example1}Full DAG}

@@

\includegraphics[width=0.4\textwidth,height=\textheight]{images/figures/dag_example2.png}

@@

\subcaption{\label{fig-dag_example2}Simplified DAG}

@@

\subsection{Additional definitions}\label{sec-appA}

\subsection{Why do we need to estimate judges'
abilities?}\label{sec-appB}

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system with plausible mechanism},
  pdfauthor={Jose Manuel Rivera Espejo; Tine van van Daal; Sven De De Maeyer; Steven Gillis},
  pdfkeywords={causal inference, probability, Thurstone, comparative
judgement, directed acyclic graph, structural causal models, statistical
modeling},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{abstract}
Dichotomous Comparative Judgment (DCJ) requires judges to compare pairs
of stimuli to determine which one exhibits a higher degree of a specific
trait. DCJ has proven effective and reliable across various fields
\citep{Pollitt_2012b, Jones_2015, vanDaal_et_al_2016, Bartholomew_et_al_2018, Lesterhuis_2018, Bartholomew_et_al_2020, Marshall_et_al_2020, Boonen_et_al_2020}.
However, despite the method's widespread use, existing literature lacks
a clear explanation of the complexities and assumptions underpinning the
DCJ system, as well as the plausible mechanisms through which DCJ data
could be generated. This study addresses these issues by representing
DCJ within the framework of causal inference. Specifically, utilizing
the structural approach, the study develops a scientific model to
clarify plausible causal assumptions and mechanisms inherent in the DCJ
system. The study then translates this model into a probabilistic
statistical model to estimate statistical relationships and infer causal
effects within the system. This research provides a robust probabilistic
foundation for the statistical analysis of DCJ data, building upon
Thurstone's law of comparative judgment \citeyearpar{Thurstone_1927}.
Its findings offer valuable insights for researchers and analysts
designing and implementing DCJ experiments.
\end{abstract}

\begin{keyword}
    causal inference \sep probability \sep Thurstone \sep comparative
judgement \sep directed acyclic graph \sep structural causal
models \sep 
    statistical modeling
\end{keyword}
\end{frontmatter}

@@

This section introduces fundamental concepts in causal inference, such
as directed acyclic graphs (DAGs), structural causal models (SCMs), and
the flow of association and causation in graphs. It is not a
comprehensive description of causal inference methods. Readers
interested in deeper exploration should consult introductory papers like
\citet{Pearl_2010}, \citet{Rohrer_2018}, \citet{Pearl_2019}, and
\citet{Cinelli_et_al_2020}. They may also find introductory books such
as \citet{Pearl_et_al_2018}, \citet{Neal_2020} and
\citet{McElreath_2020} useful. For more advanced study, seminal
intermediate papers like \citet{Neyman_et_al_1923}, \citet{Rubin_1974},
\citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009}, as well as books
such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014} and
\citet{Hernan_et_al_2020} are recommended.

@@

One key advantage of DAGs is that, regardless of complexity, they can
represent various causal structures using only five fundamental building
blocks \citep{Neal_2020, McElreath_2020}. Figure~\ref{fig-dags}
illustrates these building blocks. Figure~\ref{fig-dag_bb1} depicts two
unconnected nodes, representing an scenario where variables \(X_{1}\)
and \(X_{2}\) are not causally related. Figure~\ref{fig-dag_bb2} shows
two connected nodes, illustrating a scenario where a parent node
\(X_{1}\) exerts a causal influence on a child node \(X_{2}\).
Consequently, \(X_{2}\) is considered a \emph{descendant} of \(X_{1}\).
Figure~\ref{fig-dag_bb3} depicts a \emph{chain} (or \emph{pipe}), where
\(X_{1}\) influences \(X_{2}\), and \(X_{2}\) influences \(X_{3}\). In
this configuration, \(X_{1}\) is a parent node of \(X_{2}\), and
\(X_{2}\) a parent node of \(X_{3}\). Furthermore, the DAG shows that
\(X_{1}\) is an \emph{ancestor} of \(X_{3}\) and that the relationship
between these variables is entirely mediated by \(X_{2}\).
Figure~\ref{fig-dag_bb4} illustrates a \emph{fork}, where variables
\(X_{1}\) and \(X_{3}\) are both influenced by \(X_{2}\). In this
scenario \(X_{2}\) is a parent node of both \(X_{1}\) and \(X_{3}\).
Finally, Figure~\ref{fig-dag_bb5} depicts a \emph{collider}, also known
as \emph{inmorality}, where variables \(X_{1}\) and \(X_{3}\) are
concurrent causes of \(X_{2}\). In this configuration, \(X_{1}\) and
\(X_{3}\) are not causally related to each other but both influence
\(X_{2}\).

@@

Given the heuristic nature of DAGs, a motivating example can illustrate
how these five fundamental building blocks can be used to construct the
causal structure of a system. This example can also help clarify
additional rules and conventions for drawing DAGs.

Consider a research problem where the causal effect of a variable \(T\)
on an outcome \(Y\) needs to be investigated. Additionally, the problem
indicates that a variable \(X\) potentially influences both \(Y\) and
\(T\). Such scenarios are not hard to imagine. For instance, \(T\) could
describe different treatments potentially influencing the binary onset
of cancer (\(Y\)) while \(X\) could denote the age of a patient. Using
an example from DCJ, {we need an example}. Figure~\ref{fig-example}
presents two graphs describing the plausible causal structure of this
system, the \emph{simplified} and the \emph{magnified} DAG.

@@

\includegraphics[width=0.45\textwidth,height=\textheight]{images/figures/dag_example1.png}

@@

\subcaption{\label{fig-dag_example1}Simplied DAG}

@@

\includegraphics[width=0.85\textwidth,height=\textheight]{images/figures/dag_example2.png}

@@

\subcaption{\label{fig-dag_example2}Magnified DAG}

@@

\subsection{Why do we need to estimate judges'
abilities?}\label{sec-appA}

\subsection{Latent variables as a mean of imputation}\label{sec-appB}

\subsection{Other comparative scenarios}\label{sec-appC}
